ReactionBarrierAnalysis

### 2024-11-29 Code/ReactionBarrierAnalysis.Rmd
Updated to include women's analysis

#Only difference between these two dataframes is the filter removing 2022
dhm_no2022 <- times |>
  filter(Type == "F" | Type == "S") |>
  filter(Gender == "M") |>
  filter(Time != "DNS") |>
  filter(RxnTime > 0) |>
  filter(Event == "100 Dash" | Event == "110 Hurdles") |>
  filter(Venue != 2022) |>
  mutate(Venue = as.factor(Venue), Heat = as.factor(Heat))
```

@@

We can explore some information about this model and replicate the results
presented in Table 2 of the paper by running the code below.
```{r}
summary(gg3b)
cat("sigma_v or std dev of venue effect: ", gg3b$mu.coefSmo[[1]]$sigb, "\n")
cat("sigma_h or std dev of heat effect: ", gg3b$sigma.coefSmo[[1]]$sigb[1], "\n")
```

And now for the model excluding 2022
```{r}
summary(gg3b)
cat("sigma_v or std dev of venue effect: ", gg3b_no2022$mu.coefSmo[[1]]$sigb, "\n")
cat("sigma_h or std dev of heat effect: ", gg3b_no2022$sigma.coefSmo[[1]]$sigb[1], "\n")
```

@@

>>>>

#Only difference between these two dataframes is the filter removing 2022
dhm_no2022 <- times |>
  filter(Type == "F" | Type == "S") |>
  filter(Gender == "M") |>
  filter(Time != "DNS") |>
  filter(RxnTime > 0) |>
  filter(Event == "100 Dash" | Event == "110 Hurdles") |>
  filter(Venue != 2022) |>
  mutate(Venue = as.factor(Venue), Heat = as.factor(Heat))
```
Let's visualize the data.
```{r}
boxplot_df <- dhm |>
  mutate(Venue = as.numeric(as.character(Venue)))
library(ggplot2)
ggplot(boxplot_df, aes(x = Venue, y = RxnTime, group = Venue)) +
  geom_boxplot() +
  labs(x = "Year", y = "Reaction Time (s)") +
       scale_y_continuous(limits = c(0.05, 0.275),
                            breaks = seq(0.05, 0.275, 0.025)) +
  expand_limits(y = 0.1) + scale_x_continuous(limits = c(1998,2024),
                                              breaks = seq(1999, 2024, 2))
```
We see that the 2022 boxplot is lower than any other year.  What is particularly
interesting is that Allen's 0.099 reaction time that led to his disqualification
does not appear as an outlier in 2022, it is just the minimum value.  Compared
to in 1999, where a value right around 0.100 was an outlier.

@@

We can explore some information about this model and replicate the results
presented in Table 2 of the paper by running the code below.
```{r}
summary(gg3b)
cat("tau_v or std dev of venue effect: ", gg3b$mu.coefSmo[[1]]$sigb, "\n")
cat("tau_h or std dev of heat effect: ", gg3b$sigma.coefSmo[[1]]$sigb[1], "\n")
```

And now for the model excluding 2022
```{r}
summary(gg3b)
cat("tau_v or std dev of venue effect: ", gg3b_no2022$mu.coefSmo[[1]]$sigb, "\n")
cat("tau_h or std dev of heat effect: ", gg3b_no2022$sigma.coefSmo[[1]]$sigb[1], "\n")
```

@@

By pooling together all events within the year comparison, we find highly
significant results, suggesting that on average, the athletes who competed in
2022 had faster reaction times.

### Women's Gamlss
We decided to separate women's results after numerous papers suggested that
men and women have differing reaction times, but we still thought it was important
to investigate what an appropriate reaction time barrier was for women.  The
women's data is very similar to the men's, but there is not data from 1999
available for women and the women compete in the 100 meter hurdle instead of the
110 meter hurdle.  We are going to replicate many of the same methods used above
for the men's results.

We read in the women's data by filtering `Gender` and `Event` differently.
```{r}
library(tidyverse)
dhw <- times |>
  filter(Type == "F" | Type == "S") |>
  filter(Gender == "F") |>
  filter(Time != "DNS") |>
  filter(RxnTime > 0) |>
  filter(Event == "100 Hurdles" | Event == "100 Dash")
dhw <- dhw[- which.min(dhw$RxnTime), ] #We remove the minimum time as it is an
#extreme outlier as we want an accurate prediction of the left tail
```

Lets visualize the women's data.  This appears as Figure 1 in the supplemental
file.
```{r}
library(ggplot2)
ggplot(dhw, aes(x = Venue, y = RxnTime, group = Venue)) +
  geom_boxplot() +
  labs(x = "Year", y = "Reaction Time (s)") +
       scale_y_continuous(limits = c(0.05, 0.3),
                            breaks = seq(0.05, 0.3, 0.025)) +
  expand_limits(y = 0.1) + scale_x_continuous(limits = c(2000,2024),
                                              breaks = seq(2001, 2024, 2))
```
We see that 2022 appears as a low outlier, but perhaps not as extreme as for the
men's data.  The women's data also seems to be more stable than the men's with
fewer venue to venue differences.

Now we are ready to fit the women's model.  We tested several types of generalized
Gamma models, including those different from the men's model, and found that
this model, which is the female equivalent to the men's best model, was the best
fit.
```{r}
dhw <- dhw|>
  mutate(Venue = as.factor(Venue), Heat = as.factor(Heat))
gg3b0.w <- gamlss(RxnTime ~ random(Venue), sigma.formula = ~ random(Heat),
                  data = dhw, family = GG, control = gamlss.control(n.cyc = 40))
```

```{r}
summary(gg3b0.w)
cat("tau_v or std dev of venue effect: ", gg3b0.w$mu.coefSmo[[1]]$sigb, "\n")
cat("tau_h or std dev of heat effect: ", gg3b0.w$sigma.coefSmo[[1]]$sigb[1], "\n")
```
We see that the heat and venue effects are statistically significant and
contribute positively to the model.  We are also going to look at the residuals
to see how well the data is fit.
```{r}
library(qqconf)
qq_conf_plot(residuals(gg3b0.w))
```
Although this fit is not as strong as the men's data, we still found it to be
acceptable and it was the best of the several women's models that we looked at.

As the structure of the model remains the same, we are able to reuse the `simfit`
code developed above in order to examine what the women's disqualification
barrier should look like.

```{r}
samplefit_w <- simfit(gg3b0.w)

mean(samplefit_w < 0.10)
cat("\n")
mean(samplefit_w < 0.09)
cat("\n")
mean(samplefit_w < 0.08)

```{r}
quantile(samplefit_w, c(.01, .001, .0001))
```

It is worth noting that these suggested times are slower than the one's for the
men's data.  Thus, men and women are not currently held to the same standard
if men are more likely to react quickly than women and if they share the 0.1
barrier.
